An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-12-03 paper/4.xpg.tex
undefined -> \it{undefined} in Section 4, this follows Section 2
update sec.4: tiny

    \noindent $\tabs{2} \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{\text{undefined}}, h \ s) \\
        \tab \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{\text{undefined}} \\
        \tab \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{\text{undefined}}$

@@

Constructions of $pg$ for the replacement, the product and the source/view rearrangements are very clear when just paring $put$ and $get$ respectively, then doing basic changes.

@@

            The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
            %We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
            Since there is no real view, this $dummy$ is necessary. The last equality changes two. First is a change from $dummy$ to an application $construct\_dummy \ s$. This shows that we need a way to construct $dummy$ from the current source, $s$. We can require programmers to give this $construct\_dummy$ function. Second is a substitution on $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. The replacement of $s$ and $s_1$ is possible because of the \textsc{PutPut} law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

When evaluating $\pginline{bx_1 \circ bx_2}$, there are three $pg$ calls, of which twice for $\pginline{bx_1}$ and once for $\pginline{bx_2}$. If a given program is a left-associative composition, the number of $pg$ calls will be exponential. Therefore, the runtime inefficiency is inevitable. To solve that, we introduce a new function, $cpg$, accumulates updates in the source and the view. $\cpg{bx}{ks}{kv}{s}{v}$ is an extension of $\pg{bx}{s}{v}$ where $ks$ and $kv$ are continuations used to hold the modification information, and $s$ and $v$ are used to keep evaluated values same as $pg$. The output of this function is a 4-tuple $(ks, kv, s, v)$. To be more convenient for presenting the definition of $cpg$ as well as the other functions later, we provide some following utility functions:\\ $fst = \lambda (x_1,x_2). x_1 , \ snd = \lambda (x_1,x_2). x_2 , \ con = \lambda ks_1. \lambda ks_2. \lambda x. ((ks_1 \ x),(ks_2 \ x))$

@@

    \noindent $\cpg{Skip \ h}{ks}{kv}{s}{v} = \text{ if } h \ s = v \text{ then } (ks, kv, s, v) \text{ else undefined}$

@@

In the places where third and/or fourth argument ($s$ and $v$) are updated by applications, the computations are also accumulated in $ks$ and/or $kv$.
%Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.
Thanks to this accumulation, in $\cpginline{bx_1 \circ bx_2}$, there are only two $cpg$ calls. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Since there is no real view here, we need to construct a dummy from the source same as $pg$. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.

@@

    \noindent $\kpg{Skip \ h}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
        \tab \text{if} \ h \ s = v \ \text{then} \ (ks, kv, ks', kv', s, v) \ \text{else undefined}$

@@

In $kpg$, basically, functions for the updates will be kept in $ks'$ and $kv'$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $ks' s$ and $kv' v$ and update $ks'$ and $kv'$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $ks_1'$ and $s_1$ in the result of the first assignment are still redundant but they are not evaluated.

>>>>

    \noindent $\tabs{2} \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{undefined}, h \ s) \\
        \tab \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{undefined} \\
        \tab \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{undefined}$

@@

Constructions of $pg$ for the replacement, the product and the source/view rearrangements are very clear when just paring $put$ and $get$ respectively, then doing basic changes.\\

@@

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
%We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
Since there is no real view, this $dummy$ is necessary. The last equality changes two. First is a change from $dummy$ to an application $construct\_dummy \ s$. This means we need a way to construct $dummy$ from the current source, $s$. We can require programmers to give this $construct\_dummy$ function. Second is a substitution under the \textsc{PutPut} law on $s$ of $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we realize that there is no loss information when computing a composition.

@@

When evaluating $\pginline{bx_1 \circ bx_2}$, there are three $pg$ calls, of which twice for $\pginline{bx_1}$ and once for $\pginline{bx_2}$. If a given program is a left associative composition, the number of $pg$ calls will be exponential. Therefore, the runtime inefficiency is inevitable. To solve that, we introduce a new function, $cpg$, accumulates updates on the source and the view. $\cpg{bx}{ks}{kv}{s}{v}$ is an extension of $\pg{bx}{s}{v}$ where $ks$ and $kv$ are continuations used to hold the modification information, and $s$ and $v$ are used to keep evaluated values same as $pg$. The output of this function is a 4-tuple $(ks, kv, s, v)$. To be more convenient for presenting the definition of $cpg$ as well as the other functions later, we provide some following utility functions:\\ $fst = \lambda (x_1,x_2). x_1 , \ snd = \lambda (x_1,x_2). x_2 , \ con = \lambda ks_1. \lambda ks_2. \lambda x. ((ks_1 \ x),(ks_2 \ x))$

@@

    \noindent $\cpg{Skip \ h}{ks}{kv}{s}{v} = \text{ if } h \ s = v \text{ then } (ks, kv, s, v) \text{ else } undefined$

@@

In the places where third and/or fourth argument ($s$ and $v$) are updated by applications, the computations are also accumulated in $ks$ and/or $kv$.
%Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.
Thanks to these accumulations, there are only two $cpg$ calls in $\cpginline{bx_1 \circ bx_2}$. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Since there is no real view here, we need to construct a dummy from the source same as $pg$. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.

@@

    \noindent $\kpg{Skip \ h}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
        \tab \text{if} \ h \ s = v \ \text{then} \ (ks, kv, ks', kv', s, v) \ \text{else } undefined$

@@

In $kpg$, basically, functions for the updates will be kept in $ks'$ and $kv'$.
$\kpginline{Skip}$ and $\kpginline{Case}$ are special cases, because they require the actual values for evaluation. Therefore we evaluate the values by application of $ks' \ s$ and $kv' \ v$ and update $ks'$ and $kv'$ with the identity functions.
%Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$.
When evaluating $\kpginline{bx_1 \circ bx_2}$, $ks_1'$ and $s_1$ in the result of the first assignment are still redundant but they are not evaluated.
